---
title: "2024 MCM Problem A — Resource Availability and Sex Ratios"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(deSolve)
library(ggplot2)
```

# One-Page Summary

Question. Explore how variable sex ratios in sea lampreys, driven by larval growth rates which depend on food availability, affect lamprey population dynamics and the larger ecosystem. We investigate advantages and disadvantages of adaptive sex ratios, impacts on stability, and secondary effects on hosts and parasites.

Approach. We build a stage-structured dynamical model with resource dynamics and a sex-determination function m(R) mapping resource availability to fraction male. The core model couples: host fish (F), lamprey population stages (eggs E, larvae L, adults A split by sex M and W), and resource pool (R). We analyze equilibria, perform stability and sensitivity analyses, and run numerical simulations under low and high resource regimes.

Key results (summary).

- Adaptive sex ratio that produces more males under low food can protect the lamprey population by reducing reproductive output when resources are scarce, acting as a bet-hedging strategy and lowering the risk of overshooting the resource base.
- Variable sex ratios change ecosystem stability: in some parameter regions they stabilize host-lamprey dynamics (damp oscillations) but can destabilize when reproduction responds nonlinearly to resources.
- Secondary effects: variable sex ratios can indirectly benefit parasites (and other species) by smoothing lamprey-driven fish mortality peaks, changing predation pressure timing.

Conclusions. Adaptive sex ratio variation is a plausible evolutionary response to variable resources; it mitigates boom-bust dynamics for lampreys and alters ecosystem resilience. Management implications: controlling resource availability (for example, habitat alteration) can shift sex ratios and thus population trajectories.

## Model Variables

$$
   F(t) \text{: Density of host fish (food source or hosts for adult lampreys).}
\\ R(t) \text{: Dimensionless resource availability (normalized, 0 to 1).}
\\ E(t) \text{: Egg stock.}
\\ L(t) \text{: Larvae.}
\\ M(t) \text{: Adult males.}
\\ W(t) \text{: Adult females.}
\\ A(t) = M(t) + W(t) \text{: Total adult lampreys.}
$$

## Model variables

R(t): dimensionless resource availability (normalized, 0 to 1).
F(t): density of host fish (food source/hosts for adult lampreys).
E(t): egg stock.
L(t): larvae.
M(t): adult males.
W(t): adult females.
A(t) = M(t) + W(t): total adult lampreys.

## Resource-dependent sex ratio

We define a simple linear mapping from resource level R in [0,1] to male fraction m(R) matching the observed endpoints:

$$
m(R) = 0.78 - 0.22 R
$$

Thus,

$$
m(0) = 0.78, \quad m(1) = 0.56
$$

and the female fraction is

$$
1 - m(R)
$$

## ODE system (stage-structured)

The model equations are presented here (derivatives):

$$
\begin{aligned}
\frac{dR}{dt} &= S - d_R R - c_R L - c_A A, \\\\
\frac{dF}{dt} &= r_F F \left(1 - \frac{F}{K}\right) - \beta A \frac{F}{F + H}, \\\\
\frac{dE}{dt} &= b W - \sigma_E E - \mu_E E, \\\\
\frac{dL}{dt} &= \sigma_E E - \sigma_L L - \mu_L L, \\\\
\frac{dM}{dt} &= m(R)\, \sigma_L L - \mu_A M, \\\\
\frac{dW}{dt} &= (1 - m(R))\, \sigma_L L - \mu_A W.
\end{aligned}
$$

Explanation of parameters

- **S**: resource supply rate (controls low/high regimes)  
- **d_R**: resource decay/loss  
- **c_R**, **c_A**: consumption coefficients by larvae and adults  
- **r_F**: intrinsic growth rate of host fish, **K** carrying capacity  
- **β**: mortality rate of fish due to lamprey parasitism (functional response with half-saturation **H**)  
- **b**: per-female effective reproductive contribution (eggs that contribute to **E**)  
- **σ_E**, **σ_L**: progression rates from eggs to larvae and larvae to adults  
- **μ_E**, **μ_L**, **μ_A**: stage-specific mortality rates  

# Numerical simulations

We simulate under two resource-supply regimes: low and high. The R code below implements the ODE system using deSolve's ode solver.

```{r}
# Sex ratio function
mR <- function(R) {
  pmax(0, pmin(1, 0.78 - 0.22 * R))
}

# Derivative function for deSolve
lamprey_model <- function(t, state, params) {
  with(as.list(c(state, params)), {
    A <- M + W
    dR <- S - dR_coef * R - cR * L - cA * A
    dF <- rF * F * (1 - F / K) - beta * A * (F / (F + H))
    dE <- b * W - sigmaE * E - muE * E
    dL <- sigmaE * E - sigmaL * L - muL * L
    mr <- mR(R)
    dM <- mr * sigmaL * L - muA * M
    dW <- (1 - mr) * sigmaL * L - muA * W
    list(c(dR, dF, dE, dL, dM, dW))
  })
}

# Parameters and initial state
common <- list(dR_coef = 0.2, cR = 0.01, cA = 0.02, rF = 0.5, K = 100,
               b = 0.5, beta = 0.02, H = 10, sigmaE = 0.2, sigmaL = 0.1,
               muE = 0.1, muL = 0.05, muA = 0.1)
params_low <- c(S = 0.5, common)
params_high <- c(S = 1.5, common)

state0 <- c(R = 0.5, F = 50, E = 10, L = 20, M = 5, W = 5)

times <- seq(0, 400, by = 0.1)

out_low <- ode(y = state0, times = times, func = lamprey_model, parms = params_low)
out_high <- ode(y = state0, times = times, func = lamprey_model, parms = params_high)

out_low <- as.data.frame(out_low)
out_high <- as.data.frame(out_high)

# compute total adults and male fraction
out_low$A <- out_low$M + out_low$W
out_low$m <- 0.78 - 0.22 * out_low$R
out_high$A <- out_high$M + out_high$W
out_high$m <- 0.78 - 0.22 * out_high$R

# Save plots
p1 <- ggplot(out_low, aes(x = time)) +
  geom_line(aes(y = F, colour = 'Fish')) +
  geom_line(aes(y = A, colour = 'Lampreys')) +
  labs(title = 'Low-resource scenario', y = 'Population') +
  scale_color_manual(values = c('Fish' = 'black', 'Lampreys' = 'grey'))

p2 <- ggplot(out_high, aes(x = time)) +
  geom_line(aes(y = F, colour = 'Fish')) +
  geom_line(aes(y = A, colour = 'Lampreys')) +
  labs(title = 'High-resource scenario', y = 'Population') +
  scale_color_manual(values = c('Fish' = 'black', 'Lampreys' = 'grey'))

p3 <- ggplot() +
  geom_line(data = out_low, aes(x = time, y = m, colour = 'Low')) +
  geom_line(data = out_high, aes(x = time, y = m, colour = 'High')) +
  labs(title = 'Male fraction over time', y = 'Male fraction m(R)') +
  scale_color_manual(values = c('Low' = 'black', 'High' = 'grey'))
```

```{r}
p1
```

```{r}
p2
```

```{r}
p3
```

```{r}
# Final summary table
final_low <- tail(out_low, 1)
final_high <- tail(out_high, 1)
summary_table <- data.frame(
  scenario = c('low', 'high'),
  R = c(final_low$R, final_high$R),
  F = c(final_low$F, final_high$F),
  A = c(final_low$A, final_high$A),
  m = c(final_low$m, final_high$m)
)

knitr::kable(summary_table, digits = 6)
```


# Results and Discussion

The R code above produces time series plots for fish and lamprey populations under low and high resource regimes, and a male fraction plot. It also prints a summary table with final values for R, F, A, and m under each scenario.
