---
title: "Multiple Linear Regression"
execute:
  echo: false
format:
  html:
    theme: flatly
    code-fold: true
    code-tools: true
---

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(mosaic)
library(ggthemes)
library(pander)
library(car)

data <- read.csv("C://Users/Alex/Documents/GitHub/Portfolio/AB_TESTS/cars.csv")
```

## Background

Building on the simple regression of car price against production year, we now ask a richer question. A car's value isn't shaped by a single feature. Three things come to mind immediately for any used-car shopper: how new it is (year produced), how much it has been driven (odometer), and how big its engine is (engine capacity). Using the same Belarus 2019 Kaggle dataset, we build a multiple linear regression to see how these three predictors jointly explain price.

## Hypothesis

We want to know whether the group of predictors, taken together, explains variation in price, and which individual predictors matter after accounting for the others.

$$
H_0: \beta_1 = \beta_2 = \beta_3 = 0 \quad(\text{No predictor is linearly related to price})
$$
$$
H_a: \text{At least one } \beta_j \neq 0
$$
$$
\alpha = 0.05
$$

## Analysis

As before, we restrict to realistic listings (produced from 1990 onward, priced under \$50,000) and drop rows with missing engine information.

```{r}
data <- data %>%
  filter(year_produced >= 1990, price_usd < 50000, !is.na(engine_capacity))
```

We fit a model predicting price from production year, odometer reading, and engine capacity:

```{r}
lm.multi <- lm(price_usd ~ year_produced + odometer_value + engine_capacity, data = data)
pander(summary(lm.multi))
```

The overall F-test at the bottom of the model summary produces a p-value far below α = 0.05, so we reject the null hypothesis and conclude that at least one predictor is linearly related to price. Looking at the individual coefficients:

- Production year has a positive coefficient, confirming newer cars cost more.
- Odometer reading has a negative coefficient, matching the intuition that more mileage lowers value.
- Engine capacity has a positive coefficient, suggesting larger engines are associated with higher prices.

Each of these predictors is significant even after adjusting for the others.

### Checking for Multicollinearity

Multiple regression assumes the predictors aren't too strongly correlated with each other. We check the Variance Inflation Factors (VIF); values below 5 are generally considered acceptable.

```{r}
pander(vif(lm.multi))
```

All VIF values are low, so multicollinearity is not a concern and each coefficient can be interpreted on its own.

### Checking the Requirements

```{r, fig.width=10, fig.height=5}
par(mfrow = c(1, 2))
plot(lm.multi, which = 1:2, pch = 16)
```

As with the simple regression, the residuals show mild right-skew and growing spread at higher fitted prices, a natural feature of price data. We proceed while noting this limitation.

## Interpretation

Together, production year, odometer reading, and engine capacity form a significant model of used-car price. Newer cars and larger engines raise the expected price, while higher mileage lowers it, exactly as a buyer would expect. Because the overall F-test and each individual predictor are significant at α = 0.05, we reject the null hypothesis and conclude that these features jointly and individually help explain a car's selling price.

## Credit/References

This analysis used mild assistance from ChatGPT for plotting and diagnostics, and follows the format of my other portfolio analyses.

[Kaggle Car Dataset](https://www.kaggle.com/datasets/lepchenkov/usedcarscatalog/data)
