---
title: "Paired Samples Test"
execute:
  echo: false
format:
  html:
    theme: flatly
    code-fold: true
    code-tools: true
---

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(mosaic)
library(car)
library(pander)
library(ggthemes)

HSS <- read_csv("C://Users/Alex/Documents/GitHub/Portfolio/AB_TESTS/HighSchoolSeniors.csv")
```

## Background

The US Census at High Schools 2020 survey asked each student two related questions: how many text messages they sent yesterday, and how many they received yesterday. Because both numbers come from the same student, they form a paired measurement. This lets us ask a more personal question than comparing two separate groups: within a single person, does the number of texts they send differ from the number they receive? If students mostly reply to conversations, sent and received counts should be close; if messages tend to arrive faster than students send them out, received counts will run higher.

## Within a student, do sent and received text counts differ?

$$
H_0: \mu_d = 0 \quad(\text{Mean difference between sent and received is zero})
$$
$$
H_a: \mu_d \neq 0 \quad(\text{Sent and received counts differ on average})
$$
$$
\alpha = 0.05
$$

Here $\mu_d$ is the mean of the per-student difference (texts sent minus texts received).

## Analysis

We keep students who answered both questions and, as in my earlier texting analysis, cap the counts below 250 to remove a few extreme outliers that would otherwise dominate.

```{r}
HSS_clean <- HSS %>%
  filter(!is.na(Text_Messages_Sent_Yesterday),
         !is.na(Text_Messages_Received_Yesterday),
         Text_Messages_Sent_Yesterday < 250,
         Text_Messages_Received_Yesterday < 250) %>%
  mutate(difference = Text_Messages_Sent_Yesterday - Text_Messages_Received_Yesterday)
```

The histogram of each student's difference (sent minus received) sits noticeably to the left of zero, our first clue that students tend to receive more texts than they send.

```{r, fig.height=4, fig.width=7, fig.align='center', message=FALSE}
ggplot(HSS_clean, aes(x = difference)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 0, color = "red", linewidth = 1) +
  labs(title = "Per-student difference (sent - received) leans negative",
       x = "Texts Sent - Texts Received",
       y = "Number of Students") +
  theme_bw()
```

```{r}
pander(favstats(~difference, data = HSS_clean))
```

A paired t-test requires the differences to be roughly normal. The Q-Q plot below checks that assumption.

```{r, fig.height=5, fig.width=5, fig.align='center'}
qqPlot(HSS_clean$difference, main = "QQ Plot of Differences",
       ylab = "Sent - Received", id = FALSE)
```

The differences show heavy tails, so we report both the paired t-test (relying on the large sample size and the Central Limit Theorem) and its non-parametric counterpart, the Wilcoxon signed-rank test, which makes no normality assumption.

```{r}
pander(t.test(HSS_clean$Text_Messages_Sent_Yesterday,
              HSS_clean$Text_Messages_Received_Yesterday,
              paired = TRUE))
```

```{r, warning=FALSE}
pander(wilcox.test(HSS_clean$Text_Messages_Sent_Yesterday,
                   HSS_clean$Text_Messages_Received_Yesterday,
                   paired = TRUE, exact = FALSE))
```

Both tests produce a p-value far below our significance level of α = 0.05, so we reject the null hypothesis. There is strong evidence that students do not send and receive the same number of texts.

## Interpretation

Texting among these high schoolers is lopsided: students receive noticeably more messages than they send. On average a student sent about 40 texts but received about 50, a gap of roughly 10 messages per day, and that difference is highly significant under both the paired t-test and the Wilcoxon signed-rank test (both p-values far below α = 0.05). We therefore reject the null hypothesis and conclude that incoming texts outpace outgoing ones. This dovetails nicely with my earlier High School Seniors t-test, where the volume of texts received was the variable of interest; here we see that, person by person, students' inboxes fill faster than they can reply.

## Credit/References

This analysis follows the format of my other portfolio analyses and used mild AI assistance for plotting. Survey details are available at the Census at School questionnaire.

[Census at School Questionnaire](https://ww2.amstat.org/censusatschool/pdfs/C@SQuestionnaire.pdf)
